%acknowledgement

\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

With the conclusion of this thesis another chapter of my life is coming to an
end, and there are many people who supported me in one way or another along my
path and I would like to express my sincere gratitude to all of them.

First and foremost, I have to thank my supervisor Peter for all the support and
assistance he gave me during the four years I spent as a member of his research
group. The selection of research projects and the way he directed me towards the
interesting questions that needed to be answered were both excellent. I would
also like to thank Elke for taking on co-supervision and always helping me with
any questions I had.

My time in the Schwerdtfeger group has been filled with countless great moments
all due to the people that I shared my time with. Lukas\_0, Lukas\_1 and
Jayson were always available for a competitive round of table tennis and helped
me tremendously with all kinds of scientific questions and problems. I want to
thank Lukas\_0, Jayson, Paul and Morten specifically for being great office
mates and for all the interesting and fruitful discussions we had. Furthermore,
thanks for all the conversations about ``crypto'' and stock markets to Antony
and for all the discussions about writing a good PhD thesis and life in general
to Odile. Especially Vesna needs to be mentioned here. As the backbone of our
group she always makes sure that everything runs smoothly for us.

I am very grateful for all the moral and financial support I received from my
parents, my grandparents and my brother. The certainty that there was always
someone who had my back, and who would stand by me no matter what happened
contributed immensely to the way I was able to approach this degree. Although,
there were around 18,000 km and around ten time zones separating us, Susi was
always ready for a phone call when I needed it, and I'm very grateful to have
her as a friend. I also wish to thank Emma and Victor for helping me to stay
active and for being amazing friends and climbing buddies. 

And last but not least, I would like to thank Bali for making the last couple of
months a little bit more special for me. Giving me breaks from writing,
entertaining me with pictures and messages while I was desk-bound and being so
supportive and motivating made finishing this thesis just so much easier.

